
---
title: 'Week 4 Homework'
author: "Anna Panorska"
date: "`r Sys.Date()`"
output:
  html_document: default
fontsize: 8pt
---

The following code chunk sets up the **working directory**:
```{r setup}
#setwd("/Users/Documents/ania/teaching/STAT4645/STAT4645Fall2026/LecturesF26/week3new")
```

# <span style="color: red;"> Due in Canvas September 24.</span>

# The usual instructions: 
1.	Your homework has to be typed. All graphs done in R
2.	Use full sentences to answer questions.
3.	Do not include R-code in the solutions.
4.	Do not include raw R output. Instead, use full sentences to answer questions.
5.	Submit a pdf or html file to Canvas.

### You should use this RMarkdown file to write your solutions in the spaces provided. Then, knit it to pdf or html and submit to Canvas. Please remember to make sure no code shows in the solution you submit for grading.

###  <span style="color: red;"> Each part of every problem is worth 3 points for the total of 18 points. </span>

<span style="color: red;"> The aim of this homework </span> is to study how the change in the number of simulated observations changes the results.


### Problem 1: Cauchy distribution and the LLN 
We will explore a distribution called Cauchy distribution. It is a symmetric distribution around zero (or some other center).The PDF of the standard Cauchy distribution is:
$f(x) = \frac{1}{\pi}\frac{1}{1+x^2}$

The goal is to see if the Cauchy distribution satisfies the LLN. To see this: 

**Question-1.1:** Generate a sample of size 1000 (or longer) from the standard Cauchy and plot the sample averages of cumulative samples, like we did for the Normal distribution to `see' the LLN. That is the points will have x-coordinate equal to the sample size varying from n=2 to n=1000, and y-coordinates will be averages of the first 2, 3, ... n observations in your sample. Include only the plot in the homework solution.

**Your graph goes here.**
```{r problem 1 part 1, fig.height = 6, fig.width = 8, echo=FALSE}

set.seed(123)

n <- 2000

x <- rcauchy(n)

sample_average <- cumsum(x) / (1:n)

plot(
  2:n,
  sample_average[2:n],
  type = "l",
  xlab = "Sample Size n",
  ylab = "Sample Average",
  main = "Running Sample Average: Standard Cauchy"
)

abline(h = 0, lty = 2)


```



**Question-1.2:**  Run this simulation several times and observe behavior of the sample averages. Do they converge to anything? Does the Cauchy distribution obey the LLN?

```{r Problem 1 part 2, fig.height = 8, fig.width = 10, echo=FALSE}
set.seed(123)

n <- 2000

sample_average <- replicate(4, {   # create 4 plots of the sample mean to show different dist's with different random seeds 
  x <- rcauchy(n)
  cumsum(x) / (1:n)  # create a vector of the running sample mean
})

par(
  mfrow = c(2, 2),
  oma = c(0, 0, 3, 0),
  mar = c(4, 4, 3, 1)
)

for (i in 1:4) {

  plot(
    2:n,
    sample_average[2:n, i],
    type = "l",
    main = paste("Simulation", i),
    xlab = "Sample Size n",
    ylab = "Sample Average"
  )

  abline(h = 0, lty = 2)
}

mtext(
  "Running Sample Averages: Standard Cauchy Distribution",
  outer = TRUE,
  line = 1,
  cex = 1.2
)

```

**Answer: ** Based on the 4 simulations in the graph above, $\bar{X}$ (the sample average for a Cauchy dist) does not converge to a single number.  For example,  Simulation1 shows a possible convergence of $\bar{X}$ to zero, but there is a spike near the 1750th observation. There the sample mean goes from 0 to 2, caused by a large observation of $X_{1750} \approx 3500$.  This is a large but not impossible value, given the heavy tails of the Cauchy distribution.  Note, the 4 simulations look different.  This is due to random seed changing for each graph, giving further evidence of the non-converging nature of the Cauchy distribution.  

**Question-1.3** Now lets see the distribution of the arithmetic average of Cauchy distributions. Plot (in one plot below) histograms of the sample averages from samples of size n=10 and n=100. Use B=1000 simulated samples for each sample size to get the observations from X_bar.

**Your graph goes here.**

```{r Problem 1 part 3, fig.height = 5, fig.width = 10, fig.align = "center", echo=FALSE}

sample_size <- c(10, 100)
simulation_size <- 1000

par(
  mfrow = c(1, 2),
  oma = c(0, 0, 3, 0),
  mar = c(4, 4, 3, 1), 
  las = 1
)

hist_breaks <- seq(-6, 6, by = 0.5)

for (i in 1:length(sample_size)) {
  
  n <- sample_size[i]

  # Store B = 1000 sample averages
  x_bar <- numeric(simulation_size)
  
  for (j in 1:simulation_size) {
    x_bar[j] <- mean(rcauchy(n))
  }

  x_plot <- x_bar[x_bar >= -6 & x_bar <= 6]

  hist(
    x_plot,
    probability = TRUE,
    breaks = hist_breaks,
    main = paste("n =", n),
    xlab = expression(bar(X)),
    ylab = "Density",
    xlim = c(-6, 6),
    ylim = c(0, 0.4)
  )

  curve(
    dcauchy(x),
    add = TRUE,
    col = "red",
    lwd = 2
  )
}

mtext(
  "Distribution of Cauchy Sample Averages",
  outer = TRUE,
  line = 1,
  cex = 1.4
)


```


**Question-1.4:** Do the sample averages from the Cauchy distribution behave like the sample averages from the normal distribution? Justify your answer?

**Answer: ** Unlike the simulation for a Normal(0,1) distribution, increasing the sample size for Cauchy(n) does not shrink the spread of the distribution as n increases from 10 to 100.  Although there is an arithmetic average of the Cauchy sample, and the values observations cluster towards this average (as shown in the graphs).  But since the mean of a Cauchy distribution does not exist, there is nothing for $\bar{X}$ to converge to, so the two histograms look similar for the different values of n.  


## Problem 2. An illustration of convergence in distribution.

In this problem we will look at the converge in distribution of exponential random variables with varying means. Let 

$X_n$ have the exponential distribution with mean=$2+1/n$.

Check the help file for the exponential distribution in R to see how to generate observations from an exponential with a given mean.

**Question2.1:** Generate random samples of size 1000 from exponential distributions with n = 2, n=5, and n=10.  Plot their CDFs in different colors on a graph. Include the graph as your answer to this question below.

**Your graph goes here.**

```{r convergence in law, fig.height = 6, fig.width = 7, fig.align = "center", echo=FALSE}
n_values <- c(2, 5, 10)
B <- 10000

colors <- c("green", "orange", "purple")

# Set up an empty plot
plot(
  NA,
  xlim = c(0, 10),
  ylim = c(0, 1),
  xlab = "x",
  ylab = "CDF",
  main = "Convergence in Distribution to exp(rate=1/2)"
)

for (i in 1:length(n_values)) {

  n <- n_values[i]

  x <- rexp(
    B,
    rate = 1/(1 + 1/n)
  )

  F_hat <- ecdf(x)

  curve(
    F_hat(x),
    from = 0,
    to = 10,
    add = TRUE,
    col = colors[i],
    lwd = 2
  )
}

# Limiting theoretical CDF
curve(
  pexp(x),
  from = 0,
  to = 10,
  add = TRUE,
  col = "black",
  lwd = 3,
  lty = 2
)

legend(
  "bottomright",
  legend = c(
    paste("n =", n_values),
    "exp(rate=1/2)"
  ),
  col = c(colors, "black"),
  lwd = c(rep(2, 4), 3),
  lty = c(rep(1, 4), 2),
  bty = "n"
)


mtext(
  "Convergence in Distribution to exp(rate=1/2)",
  outer = TRUE,
  cex = 1.3
)

```

**Question2.2:** Do the CDFs seem to converge to another CDF as n increases? If yes, what is the **mean** of the limit exponential distribution?

**Answer: **  The simulation generates CDF's for the exponential distribution where the **rate** parameter varies as a function of n.  Four curves are generated for the **rate**  parameter, equal to $\frac{1}{2 + \frac{1}{n}}$  As n increases the resulting curves get closed to the true CDF of **exp(rate=1/2)**





**Note: The following math is added personal future reference.**

### Discussion on Expectation of the Cauchy Distribution

For a continuous random variable \(X\), the expected value is

$$
E[X] = \int_{-\infty}^{\infty} x f_X(x)\,dx.
$$

For a standard Cauchy random variable,

$$
X \sim \operatorname{Cauchy}(0,1),
$$

the probability density function is

$$
f_X(x) = \frac{1}{\pi(1+x^2)}.
$$

Therefore, the expected value would be

$$
E[X]
=
\int_{-\infty}^{\infty}
x \frac{1}{\pi(1+x^2)}\,dx
=
\frac{1}{\pi}
\int_{-\infty}^{\infty}
\frac{x}{1+x^2}\,dx.
$$

Because this is an improper integral, we must examine the positive and negative parts separately:

$$
E[X]
=
\frac{1}{\pi}
\left[
\int_{-\infty}^{0}\frac{x}{1+x^2}\,dx
+
\int_{0}^{\infty}\frac{x}{1+x^2}\,dx
\right].
$$

### Positive Part

Consider

$$
\frac{1}{\pi}
\int_0^\infty
\frac{x}{1+x^2}\,dx.
$$

Using the substitution

$$
u = 1+x^2,
\qquad
du = 2x\,dx,
$$

we obtain

$$
\int \frac{x}{1+x^2}\,dx
=
\frac{1}{2}\ln(1+x^2).
$$

Therefore,

$$
\begin{aligned}
\frac{1}{\pi}
\int_0^\infty
\frac{x}{1+x^2}\,dx
&=
\frac{1}{2\pi}
\lim_{b\to\infty}
\left[\ln(1+x^2)\right]_0^b \\[4pt]
&=
\frac{1}{2\pi}
\lim_{b\to\infty}
\ln(1+b^2) \\[4pt]
&=
+\infty.
\end{aligned}
$$

Thus,

$$
\frac{1}{\pi}
\int_0^\infty
\frac{x}{1+x^2}\,dx
=
+\infty.
$$

Similarly, on the negative side,

$$
\frac{1}{\pi}
\int_{-\infty}^{0}
\frac{x}{1+x^2}\,dx
=
-\infty.
$$

Consequently, computing the expectation would require

$$
E[X] = -\infty + \infty,
$$

which is undefined. Therefore,

$$
\boxed{E[X] \text{ does not exist}.}
$$

### Connection to the Heavy Tails

For large \(|x|\), the Cauchy density behaves approximately as

$$
f_X(x)
=
\frac{1}{\pi(1+x^2)}
\approx
\frac{1}{\pi x^2}.
$$

When calculating the expectation, however, the density is multiplied by \(x\):

$$
x f_X(x)
\approx
x\frac{1}{\pi x^2}
=
\frac{1}{\pi x}.
$$

Thus, the positive part of the expectation behaves like

$$
\int_1^\infty \frac{1}{\pi x}\,dx.
$$

But

$$
\int_1^b \frac{1}{x}\,dx
=
\ln b,
$$

and

$$
\lim_{b\to\infty}\ln b
=
\infty.
$$

Therefore, the heavy tails of the Cauchy distribution cause the expectation integral to diverge:

$$
\boxed{
f_X(x)\sim\frac{1}{x^2}
\quad\Longrightarrow\quad
x f_X(x)\sim\frac{1}{x}
\quad\Longrightarrow\quad
\int_1^\infty\frac{1}{x}\,dx=\infty
\quad\Longrightarrow\quad
E[X]\text{ does not exist}.
}
$$

### Why Symmetry Does Not Give a Mean of Zero

The Cauchy distribution is symmetric about zero. If we force the integration limits to expand symmetrically,

$$
\lim_{a\to\infty}
\int_{-a}^{a}
\frac{x}{\pi(1+x^2)}\,dx
=
0.
$$

This is called the **Cauchy principal value**. However, the Cauchy principal value is not the same as the expected value.

For the expected value to exist, the positive and negative contributions must be well-defined separately. For the Cauchy distribution, they diverge to \(+\infty\) and \(-\infty\), respectively. Therefore, despite its symmetry,

$$
\boxed{E[X]\text{ is undefined, not }0.}
$$


