---
title: "Assignment2"
author: "Greg T."
date: "2026-09-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: Probability as Relative Frequency

We will use relative frequency to approximate the probability that an observation from a standard normal distribution falls between 1 and 2. This is similar to a simulation we did in class in Week 2.

* Generate random samples of sizes \(10\), \(50\), \(100\), and \(1000\) from the standard normal distribution.
* For each sample, compute the relative frequency of observations between \(1\) and \(2\).
* Compute the theoretical (exact) probability that an observation from a standard normal distribution falls between \(1\) and \(2\).
* What happens to the accuracy of the relative-frequency approximation of the probability as the sample size increases? Write 1–2 sentences that are concise and directly answer the question.



```{r, echo = FALSE}
sample_size <- c(10, 50, 100, 1000)

upper_limit <- 2
lower_limit <- 1

prob_slice <- numeric(length(sample_size))
count_slice <- numeric(length(sample_size))

for (i in 1:length(sample_size)) {

  n <- sample_size[i]

  # Generate n observations from N(0,1)
  x <- rnorm(n, mean = 0, sd = 1)

  # Logical vector: TRUE if observation is between 1 and 2
  in_slice <- logical(n)

  for (j in 1:n) {

    in_slice[j] <- (x[j] < upper_limit && x[j] > lower_limit)

  }

  count_slice[i] <- sum(in_slice)
  prob_slice[i]  <- mean(in_slice)

#  cat(
#    "n =", n,
#    "count within UL and LL =", count_slice[i],
#    "relative frequency =", prob_slice[i],
#    "\n"
#  )
  
 

}

 exact_prob <- pnorm(2) - pnorm(1)

  results <- data.frame(
    n = sample_size,
    count = count_slice,
    relative_frequency = prob_slice,
    exact_probability = exact_prob,
    error = prob_slice - exact_prob
  )

knitr::kable(
  results,
  caption = "Table 1: Relative Frequency Compared with Exact Probability"
)

```
### Task-1 Discussion: 
This simulation examines how the relative frequency of observations falling between 1 and 2 approaches the corresponding theoretical probability for a standard normal distribution. For each sample size \(n=10,50,100,\) and \(1000\), a random sample is generated from \(N(0,1)\). Each observation is then checked to determine whether it lies between the lower limit 1 and the upper limit 2. The number of observations in this interval is counted, and the relative frequency is calculated as the proportion of observations satisfying the condition.

The theoretical probability is computed using

$$
P(1<X<2)=\Phi(2)-\Phi(1),
$$

where \(\Phi\) is the standard normal CDF. The simulated relative frequencies are then compared with this exact probability, and the difference is recorded as the simulation error. As the sample size increases, the relative frequency should generally become closer to the theoretical probability, illustrating the Law of Large Numbers.



### Answer to Task-1 Question: 
As the size of the ramdom sample increases, the error between the simulated `relative frequency` of observations between 1 and 2 and the calculates exact probability using the R function `pnorm` decreases. Table1 shows this result in the `error` column. 


## Task 2: Histogram as an Approximation of the Probability Density Function (PDF)

We will investigate how the sample size (number of simulated observations) changes the approximation of the probability density function (PDF) using a histogram.

* Generate random samples of sizes \(10\), \(50\), \(100\), and \(1000\) from the standard normal distribution.
* Plot a probability histogram for each sample. Plot all four histograms on one plotting page (see the code from lectures).
* Add the standard normal PDF curve to each histogram. In your homework solution, include only one plot: the plot containing the histograms and PDF curves.
* A histogram is an approximation of the PDF. How does the accuracy of this approximation change as the sample size increases? Write 1–2 sentences that are concise and directly answer the question.
* Now think of probability as the area under the PDF curve. These theoretical areas (probabilities) are approximated by relative frequencies, which correspond to areas under the histogram. How does the accuracy of the relative frequencies (areas under the histogram) as approximations of probability change as the sample size increases?



```{r, echo=FALSE}
sample_size <- c(10, 50, 100, 1000)

par(
  mfrow = c(2, 2),
  oma = c(0, 0, 3, 0),
  mar = c(4, 4, 3, 1), 
  las = 1
)

# Same histogram bins for every panel
hist_breaks <- seq(-6, 6, by = 0.5)

for (i in 1:length(sample_size)) {

  n <- sample_size[i]

  # Standard normal
  x <- rnorm(n, mean = 0, sd = 1)

  hist(
    x,
    probability = TRUE,
    breaks = hist_breaks,
    main = paste("n =", n),
    xlab = "x",
    ylab = "Density",
    xlim = c(-4, 4),
    ylim = c(0, 0.7)
  )

  curve(
    dnorm(x, mean = 0, sd = 1),
    add = TRUE,
    col = "red",
    lwd = 2
  )
}

mtext(
  "Histogram Approximation of the Standard Normal PDF",
  outer = TRUE,
  side = 3,
  line = 1,
  cex = 1.4,
  font = 2
)

```

### Task-2 Discussion: 
This simulation illustrates how a histogram constructed from a random sample approaches the theoretical probability density function of the standard normal distribution as the sample size increases. Samples of size \(n=10,50,100,\) and \(1000\) are generated from \(N(0,1)\), and a probability histogram is produced for each sample. The theoretical standard normal density curve is overlaid in red for comparison.



### Answer to Task-2 Question: 
As the sample size increases, the histogram more closely matches the shape of the standard normal density curve. This demonstrates how the empirical distribution of a large random sample provide a better approximation of the underlying population distribution.
