---
title: "Assignment1"
author: "Greg T."
date: "2026-09-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Tasks for the Assignment  

- Generate a random sample of size 100 from the standard normal distribution and plot the empirical quantile function for your sample. 
- Find the true 35th and 55th quantiles of the standard normal distribution. 
- Compute the 35th and 55th empirical quantiles from your data. Are they close to the true/theoretical ones? 
- What would you do to get empirical quantiles closer to the theoretical ones?


### Assignment discussion: 

The empirical quantile function is closely related to the empirical cumulative distribution function (ECDF). In fact, it can be viewed as a generalized inverse of the ECDF.

Suppose we have a random sample

$$
X_1, X_2, \ldots, X_n.
$$

After sorting the observations, we obtain the order statistics

$$
X_{(1)} \leq X_{(2)} \leq \cdots \leq X_{(n)}.
$$

The empirical cumulative distribution function is

$$
F_n(x)
=
\frac{1}{n}
\sum_{i=1}^{n}
I(X_i \leq x),
$$

where $I(\cdot)$ is the indicator function. The ECDF gives the proportion of observations that are less than or equal to a particular value $x$.

Thus, the ECDF maps a data value to a probability:

$$
x \longmapsto F_n(x).
$$

The empirical quantile function reverses this relationship. Given a probability $p$, it returns the smallest observed value whose empirical cumulative probability is at least $p$.

Formally,

$$
Q_n(p)
=
\inf\left\{
x : F_n(x) \geq p
\right\},
\qquad 0 < p \leq 1.
$$

Therefore, the empirical quantile function maps

$$
p \longmapsto Q_n(p).
$$

For a sample of size $n$, the ECDF increases by increments of

$$
\frac{1}{n}.
$$

Consequently, the empirical quantile function can be described using the order statistics. If

$$
\frac{k-1}{n} < p \leq \frac{k}{n},
$$

then

$$
Q_n(p)=X_{(k)}.
$$

Equivalently,

$$
Q_n(p)
=
X_{(\lceil np \rceil)},
$$

where $\lceil np \rceil$ denotes the smallest integer greater than or equal to $np$.

### Example

Suppose the sorted sample is

$$
2,\;4,\;5,\;9,\;12.
$$

Here,

$$
n=5,
$$

so each observation corresponds to a probability increment of

$$
\frac{1}{5}=0.2.
$$

The empirical quantile function is therefore

$$
Q_n(p)
=
\begin{cases}
2,  & 0 < p \leq 0.2,\\
4,  & 0.2 < p \leq 0.4,\\
5,  & 0.4 < p \leq 0.6,\\
9,  & 0.6 < p \leq 0.8,\\
12, & 0.8 < p \leq 1.
\end{cases}
$$

For example,

$$
Q_n(0.35)=4
$$

because the second observation is the first value for which the ECDF reaches or exceeds $0.35$.

Similarly,

$$
Q_n(0.75)=9.
$$

### Relationship to the ECDF

The ECDF and empirical quantile function contain essentially the same information, but the axes are reversed.

For the ECDF,

$$
x \rightarrow F_n(x),
$$

so the horizontal axis contains observed data values and the vertical axis contains cumulative probabilities.

For the empirical quantile function,

$$
p \rightarrow Q_n(p),
$$

so the horizontal axis contains probabilities and the vertical axis contains observed data values.

Both functions are step functions because they are constructed from a finite sample.

### Plotting the empirical quantile function in R

The empirical quantile function can be plotted by sorting the observations and associating them with the probability levels

$$
\frac{1}{n},
\frac{2}{n},
\ldots,
\frac{n}{n}.
$$

For example:

```r
x_sorted <- sort(x)
n <- length(x)

plot(
    (1:n) / n,
    x_sorted,
    type = "s",
    xlab = "p",
    ylab = "Empirical Quantile",
    main = "Empirical Quantile Function"
)
```

The vector

```r
(1:n) / n
```

constructs the empirical probability levels

$$
\frac{1}{n},\frac{2}{n},\ldots,1,
$$

while

```r
sort(x)
```

produces

$$
X_{(1)},X_{(2)},\ldots,X_{(n)}.
$$

Thus each plotted point corresponds conceptually to

$$
\left(
\frac{k}{n},
X_{(k)}
\right).
$$

The step plot represents the empirical quantile function obtained from the observed sample.

### Empirical versus theoretical quantiles

If the sample comes from a distribution with cumulative distribution function $F(x)$, then the theoretical quantile function is

$$
Q(p)=F^{-1}(p).
$$

For example, for a standard normal distribution, the theoretical quantile function is

$$
Q(p)=\Phi^{-1}(p).
$$

In R, this is calculated using

```r
qnorm(p)
```

while empirical quantiles are obtained from the observed sample.

As the sample size increases, the empirical quantile function generally becomes a better approximation to the population quantile function.


```{r}

set.seed(123)

x <- rnorm(100)
n <- length(x)

plot(
    (1:n) / n,
    sort(x),
    type = "s",
    xlab = "p",
    ylab = "Empirical Quantile",
    main = "Empirical Quantile Function"
)

# overlay the Theoritical Normal Quantile function

curve(
    qnorm(x),
    from = 0.001,
    to = 0.999,
    add = TRUE,
    col = "red",
    lwd = 2
)

```


```{r}

x_sorted <- sort(x)
#cat("vector length is", n, "\n")
q_55 <- x_sorted[ceiling(0.55 * n)] #return the 55th element of the sorted vector 
cat("Emperical Q_55 = ", q_55, "\n")
cat("Analytic Q_55", qnorm(.55), "\n\n")

q_35 <- x_sorted[ceiling(0.35 * n)] # return the 35th element of the sorted vector. 
cat("Emperical Q_35 = ", q_35, "\n")
cat("Analytic Q_35", qnorm(.35), "\n")

```
```{r}

set.seed(123)

x <- rnorm(1000)
n <- length(x)

plot(
    (1:n) / n,
    sort(x),
    type = "s",
    xlab = "p",
    ylab = "Empirical Quantile",
    main = "Empirical Quantile Function"
)

# overlay the Theoritical Normal Quantile function

curve(
    qnorm(x),
    from = 0.001,
    to = 0.999,
    add = TRUE,
    col = "red",
    lwd = 2
)
```

```{r}

x_sorted <- sort(x)
# cat("vector length is", n, "\n")
q_55 <- x_sorted[ceiling(0.55 * n)]  # return the 550th element of the sorted vector
cat("Emperical Q_55 = ", q_55, "\n")
cat("Analytic Q_55", qnorm(.55), "\n\n")

q_35 <- x_sorted[ceiling(0.35 * n)]  # return the 350th element of the sorted vector
cat("Emperical Q_35 = ", q_35, "\n")
cat("Analytic Q_35", qnorm(.35), "\n")

```











