---
title: "Week3 In Class Work "
author: "GregT"
date: "2026-09-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r}
# Number of observations
n <- 1000

# Generate iid observations from N(5, 4)
set.seed(123)
x <- rnorm(n, mean = 5, sd = 2)  # use sqrt(4), rnorm takes the standard deviation

# Compute the running sample mean
x_mean <- cumsum(x) / (1:n)  # element wise vector division

# Plot the running sample mean
plot(1:n, x_mean,
     type = "l",
     xlab = "n",
     ylab = "Runqning Sample Mean",
     main = "Convergence of Sample Mean")

# Add the true population mean
abline(h = 5, col = "red", lwd = 2)

```




```{r}
sample_size <- c(5, 10, 20, 50, 100, 200, 500, 1000)

epsilon <- 0.25
sim_size <- 10000

# One probability for each sample size
# numeric(length(sample_size)) creates a numeric vector of length 8, initialized w/zeros
# the prob_outlier vector is needed graphing (y-axis)
prob_outlier <- numeric(length(sample_size))
count_outlier <- numeric(length(sample_size))

#Loop1
# loop over the sample size vector, this is the x-axis in the graph 
for (i in 1:length(sample_size)) {

  n <- sample_size[i]  

  outlier <- logical(sim_size)  # initialize a boolean vector length sim_size
  
  #Loop2
  # Create the x vector and set the T/F condition based on exceeding epsilon 
  for (j in 1:sim_size) {

    x <- rnorm(n, mean = 5, sd = 2)
    x_bar <- mean(x)                  # could replace with mean(rnorm(n,5,2))

    outlier[j] <- abs(x_bar - 5) > epsilon
  } # end of the simulaton loop

  # Store results
  count_outlier[i] <- sum(outlier)
  prob_outlier[i] <- mean(outlier)

  cat(
    "n =", n,
    "count exceeds epsilon =", count_outlier[i],
    " outlier probability =", prob_outlier[i],
    "\n"
  )
} # end of the sample size loop 

# Create results table
results <- data.frame(
  n = sample_size,
  count_exceeds_epsilon = count_outlier,
  outlier_probability = prob_outlier
)

# Display table
knitr::kable(
  results,
  digits = 4,
  col.names = c(
    "n",
    "Count",
    "Probability"
  ),
  caption = "Convergence in Probability Simulation",
  escape = FALSE
)


# Plot results
plot(
  sample_size,
  prob_outlier,
  type = "b",
  xlab = "Sample Size (n)",
  ylab = "P(|Xbar - 5| > 0.25)",
  main = "Convergence in Probability"
)

```

### Theoretical Probability for the Simulation

Because the original data are normally distributed, there is an **exact theoretical probability** for every sample size \(n\). The simulation results can therefore be compared directly with the theoretical values.

We generate observations from

$$
X_i \sim N(5,2^2).
$$

Therefore, the sample mean has distribution

$$
\bar X \sim N\left(5,\frac{2^2}{n}\right),
$$

so the standard deviation of the sample mean, or standard error, is

$$
\sigma_{\bar X}=\frac{2}{\sqrt n}.
$$

We are interested in the probability

$$
P(|\bar X-5|>0.25).
$$

Standardizing gives

$$
P\left(
\left|
\frac{\bar X-5}{2/\sqrt n}
\right|
>
\frac{0.25}{2/\sqrt n}
\right).
$$

Since the standardized quantity is

$$
Z \sim N(0,1),
$$

we obtain

$$
P(|\bar X-5|>0.25)
=
2\left[
1-\Phi\left(\frac{0.25\sqrt n}{2}\right)
\right].
$$

For \(n=5\),

$$
\frac{0.25\sqrt 5}{2}\approx 0.2795,
$$

so

$$
P(|\bar X-5|>0.25)
=
2[1-\Phi(0.2795)]
\approx 0.7799.
$$

The simulation gave a probability of **0.7767**, which is very close to the exact theoretical probability.

The simulation results can be compared with the theoretical probabilities:

| \(n\) | Simulation | Exact Probability |
| ----: | ---------: | ----------------: |
|     5 |     0.7767 |            0.7799 |
|    10 |     0.6929 |            0.6926 |
|    20 |     0.5738 |            0.5762 |
|    50 |     0.3817 |            0.3768 |
|   100 |     0.2107 |            0.2113 |
|   200 |     0.0775 |            0.0771 |
|   500 |     0.0066 |           0.00519 |
|  1000 |     0.0000 |         0.0000772 |

The theoretical probability for \(n=1000\) is not exactly zero. It is approximately

$$
7.72\times10^{-5}.
$$

With \(10{,}000\) simulations, the expected number of simulations exceeding \(\epsilon=0.25\) is

$$
10{,}000(0.0000772)\approx0.77.
$$

Therefore, observing zero outliers in a particular set of \(10{,}000\) simulations is quite plausible.




```{r}
replicate(10, mean(rnorm(100, 5, 2)))

```


```{r}
#sample_size <- c(5, 10, 20, 50, 100, 200, 500, 1000)
sample_size <- c(5, 50, 200,1000)

par(
  mfrow = c(2, 2),
  oma = c(0, 0, 3, 0)
)


sim_size <- 10000
mean_vector <- numeric(sim_size)

for (i in 1:length(sample_size)){
  
  n = sample_size[i]
  
  for (j in 1:sim_size){
    x = rexp(n=n, rate = 1)
    x_bar = mean(x)
    mean_vector[j] <- x_bar
    
  }
  
  hist(
    mean_vector,
    probability = TRUE,
    main = paste("n =", n),
    xlab = "x",
    xlim = c(0, 2.5)
  )
  
  curve(
    dnorm(x, mean = 1, sd = 1 / sqrt(n)),
    add = TRUE,
    col = "red",
    lwd = 2
  )
  
  
}

mtext(
    "Central Limit Theorem Simulation for exponential dist. ",
    outer = TRUE,
    side = 3,
    line = 1,
    cex = 1.5,
    font = 2
  )
```


